---
title: "Titanic"
author: "Angel Mak"
date: "July 4, 2015"
output: html_document
---

# Overview

Goal :  
Predict whether Titanic passenger survived(1) or not(0)

Expected accuracy:  
"Know that a score of 0.79 - 0.81 is doing well on this challenge, and 0.81-0.82 is really going beyond the basic models!" - https://www.kaggle.com/c/titanic/details/getting-started-with-random-forests  

"most people above 0.85 are almost definitely cheating"
https://www.kaggle.com/c/titanic/forums/t/4894/what-accuracy-should-i-be-aiming-for  

Kaggle competitle homepage:  
https://www.kaggle.com/c/titanic

Data cleaning was done based on this tutorial:  
https://github.com/wehrley/wehrley.github.io/blob/master/SOUPTONUTS.md

# Import train and test data

```{r, message=FALSE, warning=FALSE}
library(caret) 
library(gdata)
library(Amelia) # for missingness plot
library(Hmisc) # bystats
set.seed(333)

train.raw <- read.csv("/Users/angelmak/Dropbox/Kaggle/Titanic/train.csv",as.is=T,na.strings=c("NA",""))
test.raw <- read.csv("/Users/angelmak/Dropbox/Kaggle/Titanic/test.csv",as.is=T,na.strings=c("NA",""))
```

# Data formatting

  What the data originally look like
```{r, message=FALSE, warning=FALSE}
head(train.raw)
summary(train.raw)
```

  A summary of data after converting variables to factor
```{r, message=FALSE, warning=FALSE}
train.clean <- train.raw
convert2Factor <- function(df,col){
  for (class in col){
    df[,class] <- factor(df[,class])
  }
  return(df)
}
factor.col <- c('Survived','Pclass','Sex','Embarked')
train.clean <- convert2Factor(train.clean,factor.col)
summary(train.clean)
```

# (Some) exploratory analysis
```{r, message=FALSE, warning=FALSE}
# Find out how much missing data and choose predictors
missmap(train.raw, main="Titanic Training Data - Missings Map", col=c("yellow", "black"), legend=FALSE)

# Understand what might influence the outcome
mosaicplot(train.clean$Pclass ~ train.clean$Survived, main="Passenger Fate by Traveling Class", shade=FALSE, color=TRUE, xlab="Pclass", ylab="Survived")
```

# Imputation of missing data

  + Impute median age by title as imputeAge
  + Impute Embarked by the most frequent value
  + Impute suspicious or missing Fare by Pclass
  + Group titles into higher level

```{r}
# Impute missing age using title
getTitle <- function(data) {
  #data <- train.raw
  #data$Name
  title.dot.start <- regexpr("\\,[A-Z ]{1,20}\\.", data$Name, TRUE)
  title.comma.end <- title.dot.start + attr(title.dot.start, "match.length")-1
  #data$Name[1]
  #title.dot.start[1]
  data$title <- substr(data$Name, title.dot.start+2, title.comma.end-1)
  data$title[which(data$title == "Ms")] <- 'Mrs'
  #data$title[which(data$title == "Mlle")] <- 'Mme'
  return (factor(data$title))
}

#Age, title, title.na
imputeMedian <- function(impute.var, filter.var, var.levels) {
  for (v in var.levels) {
    impute.var[which(filter.var == v)] <- impute(impute.var[which(filter.var == v)], what="median")
  }
  return (impute.var)
}

imputeAgeByTitle <- function(df){
  age.stat <- bystats(df$Age, df$title,fun=function(x)c(Mean=mean(x),Median=median(x)))
  title.na <- names(age.stat[age.stat[,"Missing"]>0,"N"])
  title.i <- which(title.na == "ALL")
  title.na <- title.na[-title.i]
  df$imputeAge <- imputeMedian(df$Age, df$title, title.na)
  return(df)
}

imputeEmbarked <- function(df){
  df$imputeEmbarked <- df$Embarked
  df$imputeEmbarked[which(is.na(df$imputeEmbarked))] <- 'S'
  return(df)
}

imputeFareByPclass <- function(df){
  df$Fare[ which( df$Fare == 0 )] <- NA
  df$Fare <- imputeMedian(df$Fare, df$Pclass, as.numeric(levels(df$Pclass)))
  return(df)
}

changeTitles <- function(df, old.titles, new.title) {
  for (old in old.titles) {
    df$newTitle[ which(df$newTitle == old)] <- new.title
  }
  return(df)
}

reassignTitle <- function(df){
  df$newTitle <- as.character(df$title)
  df <- changeTitles(df, c("Capt", "Col", "Don", "Dona", "Dr", "Jonkheer", "Lady", "Major", "Rev", "Sir"),"Noble")
  df <- changeTitles(df, c("the Countess", "Ms"),"Mrs")
  df <- changeTitles(df, c("Mlle", "Mme"),"Miss")
  df$newTitle <- factor(df$newTitle)
  return(df)
}


train.clean$title <- getTitle(train.clean)
table(train.clean$title)
train.clean <- reassignTitle(train.clean)
train.clean <- imputeAgeByTitle(train.clean)
train.clean[train.clean$title == "Dr",]
train.clean <- imputeEmbarked(train.clean)
#Impute Fare if 0 / missing
train.clean <- imputeFareByPclass(train.clean)
#Explore relationship between title and survival
table(train.clean$title,train.clean$Survived)
```

What does the data look like after cleaning
```{r}
summary(train.clean)
```

# Build models without imputated value except imputeEmbarked

  + Random forst
  + Boosting with tree (gbm)

```{r, message=FALSE, warning=FALSE}
# No imputateAge
train1.col <- c('Survived','Pclass','Sex','SibSp','Parch','Fare','imputeEmbarked')
train1 <- train.clean[train1.col]
train <- train1
inTrain <- createDataPartition(y=train$Survived, p=0.75,list=FALSE)
training <- train[inTrain,]
testing <- train[-inTrain,]

#train(method = "rf") throws an error when trying to fit a model that has NaN in the response variable.
ctrl <- trainControl(method = "cv",number=20)
modFit1<-train(factor(Survived)~.,data=training,method="rf",prox=TRUE,trControl=ctrl)
pred.train<-predict(modFit1,training)
cf1.train <- confusionMatrix(training$Survived,pred.train)
pred.test<-predict(modFit1,testing)
cf1.test <-confusionMatrix(testing$Survived,pred.test)

modFit1.gbm<-train(factor(Survived)~.,data=training,method="gbm",trControl=ctrl,verbose=FALSE)
pred.train.gbm<-predict(modFit1.gbm,training)
cf1.train.gbm <- confusionMatrix(training$Survived,pred.train.gbm)
pred.test.gbm<-predict(modFit1.gbm,testing)
cf1.test.gbm <-confusionMatrix(testing$Survived,pred.test.gbm)
```

# Build a model with imputated value
  + Random forst
  + Boosting with tree (gbm)

```{r, message=FALSE, warning=FALSE}
train2.col <- c('Survived','Pclass','Sex','SibSp','Parch','Fare','newTitle','imputeAge','imputeEmbarked')
train2 <- train.clean[train2.col]

train <- train2
inTrain <- createDataPartition(y=train$Survived, p=0.75,list=FALSE)
training <- train[inTrain,]
testing <- train[-inTrain,]

ctrl <- trainControl(method = "cv",number=20)
modFit2<-train(factor(Survived)~.,data=training,method="rf",prox=TRUE,trControl=ctrl)
pred.train<-predict(modFit2,training)
cf2.train <- confusionMatrix(training$Survived,pred.train)
pred.test<-predict(modFit2,testing)
cf2.test <- confusionMatrix(testing$Survived,pred.test)

modFit2.gbm<-train(factor(Survived)~.,data=training,method="gbm",trControl=ctrl,verbose=FALSE)
pred.train.gbm<-predict(modFit2.gbm,training)
#varImp(modFit2.gbm)
cf2.train.gbm <- confusionMatrix(training$Survived,pred.train.gbm)
pred.test.gbm <-predict(modFit2.gbm,testing)
cf2.test.gbm <- confusionMatrix(testing$Survived,pred.test.gbm)
```

# In sample and out of sample error of each model

data: no imputation  
model: random forest  
```{r}
as.numeric(1-cf1.train$overall["Accuracy"])
#Out of sample error
as.numeric(1-cf1.test$overall["Accuracy"])
```

data: no imputation  
model: gbm  
```{r}
as.numeric(1-cf1.train.gbm$overall["Accuracy"])
#Out of sample error
as.numeric(1-cf1.test.gbm$overall["Accuracy"])
```

data: imputation  
model: random forest  
```{r}
as.numeric(1-cf2.train$overall["Accuracy"])
#Out of sample error
as.numeric(1-cf2.test$overall["Accuracy"])
```

data: imputation  
model: gbm  
```{r,eval=FALSE}
as.numeric(1-cf2.train.gbm$overall["Accuracy"])
#Out of sample error
as.numeric(1-cf2.test.gbm$overall["Accuracy"]) #*
```

# Prediction on test data

The gbm model with imputed data has the highest accuracy and so it is applied on the test data. Imputation was done in the same way as the training data.

```{r, message=FALSE, warning=FALSE,eval=FALSE}
head(test.raw)
summary(test.raw)
test.clean <- test.raw
factor.col.test <- c('Pclass','Sex','Embarked')
test.clean <- convert2Factor(test.clean,factor.col.test)
test.clean$title <- getTitle(test.clean)
test.clean<- reassignTitle(test.clean)
table(test.clean$title)
table(test.clean$newTitle)
test.clean <- imputeAgeByTitle(test.clean)
#Impute Fare if 0 / missing
test.clean <- imputeFareByPclass(test.clean)
test.clean <- imputeEmbarked(test.clean)
test.clean$imputeAge<-as.integer(test.clean$imputeAge)
test.clean[which(is.na(test.clean$imputeAge)),]
summary(test.clean)

train2.col
test.col <- c('Pclass','Sex','SibSp','Parch','Fare','newTitle','imputeAge','imputeEmbarked')
rTest <- test.clean[test.col]
pred.rTest<-predict(modFit2.gbm,rTest)
out <- data.frame(PassengerId=test.clean$PassengerId, Survived=pred.rTest)
write.csv(out,file="/Users/angelmak/Dropbox/Kaggle/Titanic/submission01_gbm.csv",col.names=T,row.names=F,quote=F)
```

